A Sound and Complete Deep Embedding of Arbitrary Normal Modal Logics in Coq

### 2025-08-20 paper/COLA/paper.tex
Change definition ordering
Minor changes

The completeness proof rely on the fact that the set of modal logic formulas is countable. We model
this fact by two functions \mintinline{coq}{encode: formula -> nat} and \mintinline{coq}{decode: nat -> option formula}
which satisfies the following property: $\forall \varphi . (\texttt{decode(encode }\varphi\texttt{)} = \varphi)$, stated in the \mintinline{coq}{countable} axiom.

@@

\begin{minted}{coq}
 Fixpoint Delta (n: nat): theory :=
   match n with
   | 0   => G
   | S n' => Insert (decode n') (Delta n')
   end.
\end{minted}

In order to produce the $\Delta_{i}$, we need to include the $i$-th formula (built using the function \mintinline{coq}{decode})
in the theory $\Delta_{i - 1}$. Finally, the maximal set is defined as:
\begin{minted}{coq}
Definition Max: theory := UnionOf Delta.
Definition UnionOf (A: nat -> theory): theory := fun t => exists i, A i t.
\end{minted}
Function \mintinline{coq}{UnionOf} is a immediate encoding the indexed set union operator using Coq. An important property
of this construction of the maximal set is that it only increases the number of elements. This fact is proved by
the followed lemma, which has an immediate proof using the previous definitions.
\begin{minted}{coq}
Lemma subset_of_insert: forall D p, Subset D (Insert p D).
\end{minted}
Next, we show that the set \mintinline{coq}{Max} is such that either \verb|p| or \verb|~p| is an element of \verb|Max|.
\begin{minted}{coq}
Lemma max_is_maximal: forall p, Max [! p !] \/ Max [! ~p !].
\end{minted}
The next step in the proof is showing that \mintinline{coq}{Max} is consistent. In order to prove this fact,
we use some lemmas. First, we show, by case analysis on formulas, that inserting an formula into a theory
preserves consistency:
\begin{minted}{coq}
Lemma insert_preserves_consistency: forall D p, Consistent A D -> Consistent A (Insert p D).
\end{minted}
And, using the previous lemma, we prove that operation \mintinline{coq}{Delta n} preserves consistency, by induction on \verb|n|.
\begin{minted}{coq}
Lemma delta_is_consistent: Consistent A G -> forall n, Consistent A (Delta n).
\end{minted}
Next, the proof that \mintinline{coq}{Max} is consistent is a corollary of the previously stated results.
\begin{minted}{coq}
Corollary max_is_consistent: Consistent A G -> Consistent A Max.
\end{minted}
The final piece is the proof of the Lindenbaum's theorem, which states that any consistent theory has
a maximal consistent set. The proof of this fact is immediate from the previous lemmas.
\begin{minted}{coq}
Theorem lindenbaum: Consistent A G -> exists D, Consistent A D /\ Maximal D /\ Subset G D.
\end{minted}
This concludes our formalization of the existence of the maximal consistent set for modal logic theories.

>>>>

The completeness proof rely on the fact that the set of modal logic formulas is countable. We model
this fact by two functions \mintinline{coq}{encode: formula -> nat} and \mintinline{coq}{decode: nat -> option formula}
which satisfies the following property: $\forall \varphi . (\texttt{decode (encode }\varphi\texttt{)} = \varphi)$, stated in the \mintinline{coq}{countable} axiom.

@@

\begin{minted}{coq}
 Fixpoint Delta (n: nat): theory :=
   match n with
   | 0   => G
   | S m => Insert (decode m) (Delta m)
   end.
\end{minted}

In order to produce the $\Delta_{i}$, we need to include the $i$-th formula (built using the function \mintinline{coq}{decode})
in the theory $\Delta_{i - 1}$. Finally, the maximal set is defined as:
\begin{minted}{coq}
Definition UnionOf (A: nat -> theory): theory := fun t => exists i, A i t.
Definition Max: theory := UnionOf Delta.
\end{minted}
Function \mintinline{coq}{UnionOf} is a immediate encoding the indexed set union operator using Coq. An important property
of this construction of the maximal set is that it only increases the number of elements. This fact is proved by
the followed lemma, which has an immediate proof using the previous definitions.
\begin{minted}{coq}
Lemma subset_of_insert: forall D p, Subset D (Insert p D).
\end{minted}
Next, we show that the set \mintinline{coq}{Max} is such that either \verb|p| or \verb|~p| is an element of \verb|Max|.
\begin{minted}{coq}
Lemma max_is_maximal: forall p, Max [! p !] \/ Max [! ~p !].
\end{minted}
The next step in the proof is showing that \mintinline{coq}{Max} is consistent. In order to prove this fact,
we use some lemmas. First, we show, by case analysis on formulas, that inserting an formula into a theory
preserves consistency:
\begin{minted}{coq}
Lemma insert_preserves_consistency: forall D p, Consistent A D -> Consistent A (Insert p D).
\end{minted}
And, using the previous lemma, we prove that operation \mintinline{coq}{Delta n} preserves consistency, by induction on \verb|n|.
\begin{minted}{coq}
Lemma delta_is_consistent: Consistent A G -> forall n, Consistent A (Delta n).
\end{minted}
Next, the proof that \mintinline{coq}{Max} is consistent is a corollary of the previously stated results.
\begin{minted}{coq}
Corollary max_is_consistent: Consistent A G -> Consistent A Max.
\end{minted}
The final piece is the proof of the Lindenbaum's theorem, which states that any consistent theory has
a maximal consistent set. The proof of this fact is immediate from the previous lemmas.
\begin{minted}{coq}
Theorem lindenbaum: Consistent A G -> exists D, Consistent A D /\ Maximal D /\ Subset G D.
\end{minted}
This concludes our formalization of the existence of the maximal consistent set for modal logic theories.
